Seasonality in governing upper-ocean dynamics at submesoscales

### 2016-06-07 writeup/rocha_etal.tex
Significant additions to the text

% CR added this
\usepackage{color}

@@

\keypoints{\item Submesoscale (10-100 km) vertical vorticity and horizontal divergence
                 divergence undergo a strong seasonal cycle that are out-of-phase.
           \item Quasi-balanced  turbulence dominates the
                  variability in late Winter/early Spring; inertia-gravity waves
                  dominate in late Summer/early Fall.
            \item These results have implications for the diagnosis of
                  surface velocity from high-resolution altimeters.
                  }

@@

\section{The LLC model outputs}
We use the output of a latitude-longitude polar cap (LLC)
MITgcm  numerical simulation  \citep{forget_etal2015}. The output analyzed here, the LLC2160,
is the 1/24$^\circ$ --- 90-vertical-levels simulation from a hierarchy of global
forward numerical solutions on a LLC grid that were
spun up from a ECCO2 adjoint-method state estimate. The LLC2160 was forced by
tides and 6-hourly surface fluxes similarly to the LLC4320 simulation
used by \citet{rocha_etal2016} --- see their appendix D. We chose the second-higher
resolution  simulation (LLC2160) of the LLC hierarchy because it was run for longer:
the LLC2160 output spans two years from March 2011 to April 2013. The model time-step
was 45s and the model variables were saved hourly. LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. While the nominal horizontal resolution  of the
LLC2160 simulation is 1/24$^\circ$,
 horizontal wavenumber spectra suggests an effective resolution of about 20 km.
 (see figure \ref{fig3}).

A key aspect of the simulation analyzed here is that it is forced by
the 16 most significant tidal components.
Barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep{rocha_etal2016}. Hence, tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper-ocean dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}.

The region of focus is the northwest Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have
shown strong seasonality in submesoscale turbulence.
We analyze a sub-domain of the LLC2160 simulation of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a-b).

\section{Seasonality in surface velocity and near-surface stratification}

Focusing on the submesoscale variability, we present estimates of bulk quantities
associated with the lateral gradient of the horizontal velocity.
The lateral gradients are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$ and
horizontal divergence $\delta \equiv u_x + v_y$.
These quantities highlight the submesoscale structures in the flow
and hint on their dynamics.

Figures \ref{fig1}a-b show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15). Clearly, a strong seasonality in vorticity
exists: large values of fine-grained vertical vorticity are
observed in early Spring with maximum values as high as $0.9f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about 0.25. In early
Fall, the situation is the opposite: the vertical vorticity is coarse-grained,
and its local maximum and RMS is no larger than $0.3f$ and $0.1f$, respectively.
Indeed, the seasonal cycle in vorticity is very strong (Figure~\ref{fig2}):
the RMS vertical vorticity is about twice as large in late Winter/early Spring
than it is in late Summer/early Fall. This picture is consistent with
recent studies: shallow baroclinic instabilities energize the submesoscales
in late Winter, drawing from the available potential energy stored in deep mixed
layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The bulk of the vertical vorticity is associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields reduces the vertical
vorticity by less than 20$\%$ and the seasonal cycle remains strong (see
green lines in figure \ref{fig2}). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity field
with a Hanning filter with cut-off scale of 100 km reduces the RMS vorticity by
at least 40$\%$, yielding bulk statistics of velocity gradient roughly consistent
with AVISO geostrophic velocities (see lines blue and black in figure \ref{fig2}).

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows are supra-inertial and project
on scales smaller than 100 km: daily-averaging or smoothing the velocity fields
 (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Interestingly, the seasonal cycle in horizontal divergence is out-of-phase with the
seasonal cycle of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.
\begin{figure*}[ht]
\begin{center}
\hspace{-1.25cm}\includegraphics[width=.7\textwidth]{figs/fig1_1.pdf}\\
\vspace{-.125cm}
\includegraphics[width=.9\textwidth]{figs/fig1_2.pdf}
 \caption{(a) The study region with the subregion where the LLC outputs are
          analyzed. LLC 4320 (1/48$^\circ$) snapshots of surface vorticity (b and c) and a transect
          of potential density at 165$^\circ$E (d and e). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

@@

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=.9\textwidth]{figs/fig4.pdf}
 \caption{Surface (horizontal) KE (a) and SSH variance wavenumber spectra (b)
 in the1/48$^\circ$ simulation. Solid lines
 are spectra based on hourly snapshots, dashed lines are spectra based on daily-averaged
 fields.}
 \label{fig3}
 \end{center}
\end{figure*}

@@

>>>>

% CR added this
\usepackage{color}
\usepackage{amsmath}

@@

\keypoints{\item Upper-ocean submesoscale (10-100 km) turbulence and innertia-gravity waves
                 undergo strong seasonal cycles that are out-of-phase.
           \item Submesoscale turbulence dominates the horizontal  velocity and
                  sea-surface height variability in late Winter/early Spring.
            \item Inertia-gravity waves are summertime enhanced near the surface and
                  dominate the surface variability in late Summer/early Fall.
                  }

@@

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
MITgcm realstic numerical simulations. The outputs
 analyzed here, LLC2160 and LLC4320,
are the 1/24$^\circ$ and 1/48$^\circ$  --- 90-vertical-levels simulations from
a hierarchy of global
forward numerical solutions on a LLC grid   \citep{forget_etal2015} that were
spun up from a ECCO2 adjoint-method state estimate. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the then 5-month long output used by
\citet{rocha_etal2016}.

A key aspect of the simulation analyzed here is that it is forced by
the 16 most significant tidal components.
Barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep{rocha_etal2016}. Tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper-ocean dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}. The LLC2160 and LLC4320
simulations fully resolve internal tides and
other hydrostatic internal waves and other submesoscale processes at scales as
small as about 5 km.

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have suggested strong submesoscale seasonality.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a strong seasonal cycle: wintertime
enhance small-scale turbulence de-stratify the upper ocean, yielding mixed-layers
as deep as 300m. In late Spring/early Summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed-layers as shallow as 40m.
The stratification and its seasonality is well-captured by both LLC simulations.
The excellent comparisons of model stratification to Argo climatology lends confidence
to the present analysis (supplemental material).

\section{Statistics of surface lateral velocity tensor}

We present estimates of bulk quantities
associated with the lateral velocity tensor
\begin{equation}
\nabla_h \bf{u}_h = \left[\begin{matrix} u_x & u_y\\ v_x&v_y \end{matrix}\right] \,.
\end{equation}
The components of the lateral velocity tensor are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$, lateral rate of strain
$\alpha \equiv [(u_x-v_y)^2 + (u_y + v_x)^2]^{1/2}$, and horizontal divergence $\delta \equiv u_x + v_y$.
These diagnostics highlight the submesoscale structures in the flow
and their analysis hint on the governing dynamics in the upper ocean
\citep{capet_etal2008a,shcherbina_etal2013}.

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions captures theseasonality in vorticity
exists: large values of fine-grained vertical vorticity are
observed in early Spring with maximum values as high as $1.5f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about 0.4. In early
Fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained,
and its local maximum and RMS are no larger than $0.5f$ and $0.2f$, respectively.
Indeed, the seasonal cycle in vorticity is very strong (figure~\ref{fig2}):
in both simulations, the RMS vertical vorticity and lateral strain rate are about twice as large in
late Winter/early Spring
than it is in late Summer/early Fall. Because the wintertime vorticity and strain
are dominated by the smallest scales in the flow, increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ significantly increases the wintertime RMS vorticity
and strain by about 40$\%$.

The bulk of vertical vorticity and lateral strain rate is associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields reduces the RMS
vorticity by less than 40$\%$ and the RMS strain by less than 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km significantly reduces the RMS
vorticity and strain. The reduction in variance is dramatic ($>$80$\%$) in winter
yielding RMS vorticity and strain roughly consistent the ones diagnosed from
AVISO gridded geostrophic velocities (compare blue lines with black lines in figures \ref{fig2}a-b).
The overall picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late Winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The seasonal cycle of the horizontal divergent part of the flow, however, add complexity
to the submesoscale-turbulence-centred scenario depicted by those invertigators.

 is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows are supra-inertial and project
on scales smaller than 100 km: daily-averaging or smoothing the velocity fields
 (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Interestingly, the seasonal cycle in horizontal divergence is out-of-phase with the
seasonal cycle of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.
\begin{figure*}[ht]
\begin{center}
\hspace{-1.25cm}\includegraphics[width=.7\textwidth]{figs/fig1_1.pdf}\\
\vspace{-.125cm}
\includegraphics[width=.9\textwidth]{figs/fig1_2.pdf}
 \caption{(a) The study region with the subregion where the LLC outputs are
          analyzed. LLC 4320 (1/48$^\circ$) snapshots of surface vorticity (b and c) and a transect
          of potential density at 165$^\circ$E (d and e). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

\section{Seasonality in governing submesoscale dynamics}

@@

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=.9\textwidth]{figs/fig4.pdf}
 \caption{Surface (horizontal) KE (a) and SSH variance wavenumber spectra (b)
 in the 1/48$^\circ$ simulation. Solid lines
 are spectra based on hourly snapshots, dashed lines are spectra based on daily-averaged
 fields.}
 \label{fig3}
 \end{center}
\end{figure*}

@@

The time-step was 45s for
the LLC2160 simulation and 25s for the LLC4320 simulation. Details of the LLC simulations
are provided in the supplemental material.

 LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. While the nominal horizontal resolution  of the
LLC2160 simulation is 1/24$^\circ$,
 horizontal wavenumber spectra suggests an effective resolution of about 20 km.
 (see figure \ref{fig3}).
